A weakly structured stem for human origins in Africa

### 2023-01-11 paper.tex
latex issue
formatting consistency

Decades of study of human genome variation have suggested a predominantly
tree-like model of recent population divergence from a single ancestral
population in Africa. It has been difficult to reconcile this finding with the
fossil and archaeological records of human occupation across the vast African
continent. For example, fossils such as those from the sites of Jebel Irhoud,
Morocco \citep{Hublin2017-cq}, Herto, Ethiopia \citep{White2003-bk} and Klasies
River, South Africa \citep{Deacon1995-rx} demonstrate that derived \emph{Homo
sapiens} anatomical features were found across the continent
300-100ka. Archaeological sites from the Middle Stone Age, of which some have
been associated with \emph{Homo sapiens}, are also widely distributed across
Africa. It is unclear whether these fossils and archaeological sites represent
populations which contributed to contemporary \emph{Homo sapiens} as population
precedents or were local ``dead-ends''. Recently, attempts to reconcile genetic
and paleoanthropological data include proposals for a Pan-African origin of
\emph{Homo sapiens} by which populations in many regions of the continent
contributed to the formation of \emph{Homo sapiens} beginning at least 300ka
\citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.

@@

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (from Sierra Leone, MSL from the
Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
Gumuz (recent descendants of a hunter-gatherer group from Ethiopia
\citep{Gurdasani2015-qy,Gopalan2022-pw}), and eastern African agriculturalists
(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
populations, despite speaking distinct Afro-Asiatic languages, are highly
genetically similar \citep{Pagani2015-pz,Gopalan2022-pw} and thus the two
groups were combined for a larger sample size (Figure~\ref{fig:diversity}). We also
included the British (GBR) from the 1000 Genomes Project in our demographic
models as a representative source of back-to-Africa gene flow and recent
colonial admixture in South Africa. Finally, we used a high-coverage ancient
Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
for gene flow from Neanderthals into non-Africans and gauge the relative
time depth of divergence, assuming Neanderthals diverged 550ka from a common
stem. We computed one- and two-locus statistics whose expectation within and across 
populations can be computed efficiently and that are well suited for both low- and
high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
maximum-likelihood inference framework, we then fit to these statistics a
family of parameterized demographic models that involve population splits, size
changes, continuous and variable migration rates, and punctuated admixture
events, to learn about the nature of population structure over the past million
years.

@@

Regardless of the model choice for early epochs, maximum-likelihood inference
of human demographic history for the last 150ka was remarkably robust.  In a
reticulated model, we use ``divergence'' between populations to mean the time
of their most recent shared ancestry. The earliest divergence among
contemporary human populations differentiates the southern African Nama from
other African groups between 110--135ka, with low to moderate levels of
subsequent gene flow (Table \ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140ka. We conclude that geographic patterns of
contemporary \emph{Homo sapiens} population structure likely arise during MIS
5. Although we find evidence for earlier population structure in Africa (next
section), contemporary populations cannot be easily mapped onto the more
ancient `stem' groups as only a small proportion of drift between contemporary
populations can be attributed to drift between stems (SI Section \ref{sec:f4},
Figures~\ref{fig:predictions} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events (SI Section \ref{sec:models}). These
parameters were ones supported by multiple genetic and archaeological studies.
Fixed parameters included the time of divergence between western and eastern
African populations, set to 60ka, just prior to the split of Eurasians and East
Africans set to 50ka. We also fixed the amount of admixture from Neanderthals
to Europeans directly following the out-of-Africa migration which was set to
1.5\% at 45ka (SI Section~\ref{sec:models}).

We quantify migration rates among populations after their divergence
$\sim$120ka. Prior to agripastoralist expansion 5ka, migration between the
ancestors of the Nama and other groups is an order of magnitude weaker than
that observed between between western and eastern Africans (Table
\ref{tab:migration-rates}). All models infer relatively high gene flow between
Ethiopia and western Africa ($m\approx2\times10^{-4}$, the constant proportion
of immigrant lineages per generation over 60ka). We further find that
Back-to-Africa gene flow at the beginning of the Holocene primarily affected
the ancestors of the Ethiopian agricultural populations
\citep{Molinaro2019-rf}, comprising over half of their genetic ancestry,
estimated to be approximately 65\%. We observe significant gene flow from the
Amhara and Oromo into the Nama, a signal which is likely a proxy for the
movement of eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 years ago. While this gene flow is not apparent from the
ADMIXTURE plot (Figure~\ref{fig:diversity}), the ancestry is likely grouped
into the Khoe-San component which has drifted appreciably from its ancestral
eastern African source. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions suggested by \texttt{ADMIXTURE}
(Figure~\ref{fig:diversity}).

@@

\begin{table}[t!]
    \centering
    \scriptsize
    \begin{tabular}{lllrrrr}
        \toprule
        \textbf{Likelihood} & \textbf{Label} & \textbf{Population Pair} &
            \textbf{Divergence} & \textbf{Migration rate} & \textbf{Migration} \\
        & & & 
            \textbf{Time (ka)} & \textbf{per generation} & \textbf{duration (kyr)} \\
        \midrule
        \textbf{Continuous Model} & & & & & \\
        $LL= -115,300$ & a & Stem 1, Stem 2 & 1,223 & 6.26e-5 & 1,089 \\
        (Table~\ref{tab:supp-continuous-migration}) & b & Stem 2, Nama & NA & 5.85e-5 & 129 \\
        & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.62e-4} & 129, 55 \\
        & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.2e-6 & 134, 60 \\
        & g & Mende, East Africans & 60 & \textbf{2.13e-4} & 60 \\
        & h & East Africans, British & 50 & 4.16e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo & 12 &\textbf{3.37e-4} & 12 \\
        \textbf{Merger Model} & & & & & \\
        $LL= -101,600$ & a & Stem 1, Stem 2 & 1,692 & \textbf{1.26e-4} & 1,213 \\
        (Table~\ref{tab:supp-merger-with-stem-migration}) & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
        & b & Stem 2 to Nama & 119 & \textbf{0.70} & pulse \\
        & c & Stem 2 to Stem 1E & 98 & \textbf{0.52} & pulse \\
        & d & Stem 2 to Mende & 25 & \textbf{0.19} & pulse \\
        & e, f & Nama, Other Africans* & 119 & 4.5e-5, 9.8e-6 & 120, 60 \\
        & g & Mende, East Africans* & 60 & \textbf{1.97e-4} & 60 \\
        & h & East Africans, British & 50 & 3.82e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo &12 & \textbf{3.59e-4} & 12 \\
        \bottomrule
    \end{tabular}
    \caption{
        \textbf{Migration and divergence parameters from best fit models.}
        Labeled migration rates correspond to symmetric continuous migration
        bands shown in Figure~\ref{fig:best-fit-models}. Both the continuous migration and
        merger models inferred a relatively deep split of human stem branches,
        though they were connected by ongoing migration that maintained their
        genetic similarity. Bold text indicates migration rates above $10^{-4}$.
        In both models, the branch ancestral to the Nama shares a common ancestral population
        with the other African groups $\sim$120--135ka. Following this divergence,
        the population ancestral to other African groups branches into West and East African
        groups 60ka.
        *Migration rates and durations are shown between branches ancestral to 
        1) Nama and East Africans and their ancestors, and
        2) Nama and Mende, respectively.
        ``Divergence times'' correspond to the most recent common ancestral population,
        and does not account for continuous migration or earlier reticulations.
    }
    \label{tab:migration-rates}
\end{table}

@@

To account for population structure prior to 135ka, three of our four models
allowed for two or more ``stem'' populations which could diverge either before
or after the Neanderthal split. We considered models both with and without
migration between these stem populations, and in both cases we tested two
different types of gene exchange during the expansion phase, as illustrated in
Figure \ref{fig:mergers}: 1) One of the stem population expands (splits into
contemporary populations), and the other stem population(s) has continuous
symmetric migration with those populations; or 2) one or more of the stem
populations expands, with instantaneous pulse (or ``merger'') events from the
other stem population, so that recent populations are formed by mergers of
multiple ancestral populations. Depending on parameter values, this scenario
encompasses archaic hominin introgression and fragmentation-and-coalescence
models (such as Figure~\ref{fig:proposed-models}B and C).  For many parameters,
confidence intervals based on bootstrapping are relatively narrow
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. Model assumptions have,
however, a larger impact on parameter estimates (and thus, real uncertainty).
To convey model uncertainty, we highlight features of the two inferred models
with high likelihoods. These are referred to as the ``multiple-merger'' and the
``continuous-migration'' models. Both allow for migration between stem
branches, but differ primarily in the timing of the early divergence of stem
populations and their relative $N_e$ (Figure~\ref{fig:best-fit-models}). The
two models also differ in the mode of divergence, with the multiple-merger
featuring a population reticulation during the Middle Pleistocene.

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration between stems ($LL \approx
-101,600$ vs.\ $-107,700$ in the merger model, Tables
\ref{tab:supp-merger-without-stem-migration} and
\ref{tab:supp-merger-with-stem-migration}, and $LL \approx -115,300$ vs.\
$-126,500$ in the continuous migration model, Tables
\ref{tab:supp-continuous-migration-no-mig} and
\ref{tab:supp-continuous-migration-no-mig}).  With continuous migration between
stems, population structure extends back to up to 1.5Ma
(Table~\ref{tab:migration-rates}).  Migration between the stems in these models
is moderate, with a fraction of migrant lineages each generation estimated as
$m=6.3\times10^{-5}-1.3\times10^{-4}$. For comparison, this is similar to
inferred migration rates between connected contemporary populations over the
past 50ka (Table~\ref{tab:migration-rates}). This ongoing (or at least,
periodic) gene flow qualitatively distinguishes these models from previously
proposed archaic hominin admixture models (Figure~\ref{fig:proposed-models}C)
as the early branches remain closely related, and each branch contributes large
amounts to all contemporary populations (Figure~\ref{fig:predictions}).
Because of this relatedness, only $1\%$ to $4\%$ of genetic differentiation
among contemporary populations can be traced back to this early population
structure (SI Section~\ref{sec:f4})

Under the continuous-migration model, one of the two stems diverges into
lineages leading to contemporary populations in western, southern and eastern
Africa, and the other (Stem 2) contributes variable ancestry to those
populations. This migration from Stem 2 is highest with the Mende
($m=1.6\times10^{-4}$) compared to the Nama and East African populations
($m=5.9\times10^{-5}$ and $3.1\times10^{-5}$, respectively), with migration
allowed to occur until 5ka. A sampled lineage from the Nama, Mende, and Gumuz
have probabilities of being in Stem 2 at the time of Stem 1 expansion (135ka)
of approximately $0.145$, $0.2$, and $0.13$, respectively, though these
probabilities change over time, precluding the notion of a fixed admixture
proportion.

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups.  We observe a
sharp bottleneck in Stem 1 down to $N_e=100$ after the split of the Neanderthal
branch. This represents the lower bound allowed in our optimization (i.e. an
$N_e$ of 100), although the size of this bottleneck is poorly constrained (95\%
confidence interval 100--851). After a long period of exchange with Stem 2,
Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S'' 478ka. The timing of
this divergence was also poorly constrained (95\% CI 276–-478ka).  These
populations evolve independently until 119ka (101--125ka) when Stem 1S and Stem
2 combine to form the ancestors of the Nama, with proportions 30\%, 70\%
respectively. Similarly, Stem 1E and Stem 2 combine in equal proportions (50\%
each) to form the ancestors of the western Africans and eastern Africans (and
thus also all individuals who later disperse during the Out of Africa event).
Finally, the Mende receive a large additional pulse of gene flow from Stem 2,
replacing 19\% (18--21\%) of their population 25ka (22--26ka). The later Stem 2
contribution to the western African Mende resulted in better model fits
($\Delta LL \approx 60,000$).  Speculatively, this may indicate that an
ancestral Stem 2 population occupied western or central Africa, broadly
speaking. The differing proportions in the Nama and eastern Africans may also
indicate geographic separation of Stem 1S in southern Africa and Stem 1E in
eastern Africa.

\begin{figure}[t!]
    \centering
    \includegraphics{figures/predictions.pdf}
    \caption{
        \textbf{Structure among stems is weak and present-day structure is mostly recent.}
        From the best fit models of our two parameterizations (A and B:
        continuous migration, C and D: merger with stem migration),
        we predicted pairwise differences $H_{i,j}$ between individuals
        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
        (B and D) To understand how drift between stems explains contemporary structure, 
        we also computed the proportion $\alpha^2$ of drift between pairs of
        sampled contemporary
        populations that aligns with drift between past populations
        (here Nama and Mende, see SI Section~S\ref{sec:f4}
        for details and additional comparisons in
        Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
        Both models infer deep population structure with modest contributions to
        contemporary genetic differentiation.
        Most present differentiation dates back to the last 100ka.
    }
    \label{fig:predictions}
\end{figure}

@@

\begin{figure}[t!]
    \centering
    \includegraphics{figures/validation}
    \caption{
        \textbf{Model validation using independent statistics.} (A--C) Using
        our best fit models to linkage disequilibrium and pairwise diversity statistics,
        we simulated expected conditional site-frequency-spectra (cSFS) and
        compared to the observed cSFS from the data. Our inferred models
        provide a good fit to the data, even though these were not used in our
        inference. Across the three populations, ancestral state
        misidentification was consistently inferred to be 1.5--1.7\% for
        intergenic loci (SI \ref{sec:misspecification}). (D, E) We used Relate
        \citep{Speidel2019-nj} to reconstruct genome-wide gene genealogies,
        which we used to estimate coalescence rate trajectories
        and cross-coalescence rates between pairs of populations. While
        coalescence rate distributions are informative statistics about past
        evolutionary processes, interpretation can be hindered by migration and
        population structure, and translating relative cross-coalescence rate
        curves (RCCR) into population divergence times is especially prone
        to misinterpretation. For example, the Mende-Gumuz comparison shows
        a more recent increased RCCR than either population with the British,
        a pattern that is recapitulated under our best-fit model, even though
        the Mende-Gumuz split occurs prior to the Gumuz-British split.
    }
    \label{fig:validation}
\end{figure}

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time in real data and data simulated from our inferred
models. Many previous studies have found a reduction of coalescence rates
between $1000$ka and $100$ka in humans, and thus inferred an increase in $N_e$
during the same period \citep{Li2011-le}.  This increase in inferred $N_e$
could be due to either an increase in population size or to ancestral
population structure during the Middle Pleistocene \citep{Mazet2016-wn}.  All
models, including the single-origin model, recapitulate an inferred ancestral
increase in $N_e$ between 100ka-1Ma extending deeper in the past than the
Relate estimate (Figure~\ref{fig:supp-iicr-sim} and SI
Section~\ref{inferred_relate}). Whereas the single-origin model achieves this by an
increase in $N_e$ during that period, the best-fit model recapitulate this
pattern without corresponding population size changes.

Relative cross-coalescence rates (rCCR) have recently been used to estimate
divergence between a pair of populations, as measured by the rate of
coalescence between two groups divided by the mean within population
coalescence. Simulations of rCCR accuracy, however, focus on a ‘clean split’
between populations whereby groups diverge without subsequent gene flow.
Published estimates of the earliest human divergences with rCCR, which range
from 150ka-100ka \citep{Bergstrom2021-iw}, may be significantly biased when
compared to more complex models with gene flow as inferred here. We find that
midpoint estimates of rCCR are poor estimates for population divergence, often
underestimating divergence time by 50\% or greater (e.g., Mende vs.\ Gumuz
$\sim$15ka compared to a true divergence of 60ka), and recent migration can
lead to the misordering of divergence events (Figure~\ref{fig:validation}E). We
suggest that rCCR analyses which do not fit multiple parameters including gene
flow should be interpreted with caution.

>>>>

Decades of study of human genome variation have suggested a predominantly
tree-like model of recent population divergence from a single ancestral
population in Africa. It has been difficult to reconcile this finding with the
fossil and archaeological records of human occupation across the vast African
continent. For example, fossils such as those from the sites of Jebel Irhoud,
Morocco \citep{Hublin2017-cq}, Herto, Ethiopia \citep{White2003-bk} and Klasies
River, South Africa \citep{Deacon1995-rx} demonstrate that derived \emph{Homo
sapiens} anatomical features were found across the continent
$300-100$ka. Archaeological sites from the Middle Stone Age, of which some have
been associated with \emph{Homo sapiens}, are also widely distributed across
Africa. It is unclear whether these fossils and archaeological sites represent
populations which contributed to contemporary \emph{Homo sapiens} as population
precedents or were local ``dead-ends''. Recently, attempts to reconcile genetic
and paleoanthropological data include proposals for a Pan-African origin of
\emph{Homo sapiens} by which populations in many regions of the continent
contributed to the formation of \emph{Homo sapiens} beginning at least $300$ka
\citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.

@@

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (from Sierra Leone, MSL from the
Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
Gumuz (recent descendants of a hunter-gatherer group from Ethiopia
\citep{Gurdasani2015-qy,Gopalan2022-pw}), and eastern African agriculturalists
(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
populations, despite speaking distinct Afro-Asiatic languages, are highly
genetically similar \citep{Pagani2015-pz,Gopalan2022-pw} and thus the two
groups were combined for a larger sample size (Figure~\ref{fig:diversity}). We also
included the British (GBR) from the 1000 Genomes Project in our demographic
models as a representative source of back-to-Africa gene flow and recent
colonial admixture in South Africa. Finally, we used a high-coverage ancient
Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
for gene flow from Neanderthals into non-Africans and gauge the relative
time depth of divergence, assuming Neanderthals diverged $550$ka from a common
stem. We computed one- and two-locus statistics whose expectation within and across 
populations can be computed efficiently and that are well suited for both low- and
high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
maximum-likelihood inference framework, we then fit to these statistics a
family of parameterized demographic models that involve population splits, size
changes, continuous and variable migration rates, and punctuated admixture
events, to learn about the nature of population structure over the past million
years.

@@

Regardless of the model choice for early epochs, maximum-likelihood inference
of human demographic history for the last $150$ka was remarkably robust.  In a
reticulated model, we use ``divergence'' between populations to mean the time
of their most recent shared ancestry. The earliest divergence among
contemporary human populations differentiates the southern African Nama from
other African groups between $110-135$ka, with low to moderate levels of
subsequent gene flow (Table \ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim140$ka. We conclude that geographic patterns of
contemporary \emph{Homo sapiens} population structure likely arise during MIS
5. Although we find evidence for earlier population structure in Africa (next
section), contemporary populations cannot be easily mapped onto the more
ancient `stem' groups as only a small proportion of drift between contemporary
populations can be attributed to drift between stems (SI Section \ref{sec:f4},
Figures~\ref{fig:predictions} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events (SI Section \ref{sec:models}). These
parameters were ones supported by multiple genetic and archaeological studies.
Fixed parameters included the time of divergence between western and eastern
African populations, set to $60$ka, just prior to the split of Eurasians and East
Africans set to $50$ka. We also fixed the amount of admixture from Neanderthals
to Europeans directly following the out-of-Africa migration which was set to
$1.5\%$ at $45$ka (SI Section~\ref{sec:models}).

We quantify migration rates among populations after their divergence
$\sim120$ka. Prior to agripastoralist expansion $5$ka, migration between the
ancestors of the Nama and other groups is an order of magnitude weaker than
that observed between between western and eastern Africans (Table
\ref{tab:migration-rates}). All models infer relatively high gene flow between
Ethiopia and western Africa ($m\approx2\times10^{-4}$, the constant proportion
of immigrant lineages per generation over $60$ka). We further find that
Back-to-Africa gene flow at the beginning of the Holocene primarily affected
the ancestors of the Ethiopian agricultural populations
\citep{Molinaro2019-rf}, comprising over half of their genetic ancestry,
estimated to be approximately $65\%$. We observe significant gene flow from the
Amhara and Oromo into the Nama, a signal which is likely a proxy for the
movement of eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a $25\%$ ancestry
contribution 2,000 years ago. While this gene flow is not apparent from the
ADMIXTURE plot (Figure~\ref{fig:diversity}), the ancestry is likely grouped
into the Khoe-San component which has drifted appreciably from its ancestral
eastern African source. Colonial period admixture from Europeans into the Nama
was estimated at $15\%$, similar to proportions suggested by \texttt{ADMIXTURE}
(Figure~\ref{fig:diversity}).

@@

\begin{table}[t!]
    \centering
    \scriptsize
    \begin{tabular}{lllrrrr}
        \toprule
        \textbf{Likelihood} & \textbf{Label} & \textbf{Population Pair} &
            \textbf{Divergence} & \textbf{Migration rate} & \textbf{Migration} \\
        & & & 
            \textbf{Time (ka)} & \textbf{per generation} & \textbf{duration (kyr)} \\
        \midrule
        \textbf{Continuous Model} & & & & & \\
        $LL= -115,300$ & a & Stem 1, Stem 2 & 1,223 & 6.26e-5 & 1,089 \\
        (Table~\ref{tab:supp-continuous-migration}) & b & Stem 2, Nama & NA & 5.85e-5 & 129 \\
        & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.62e-4} & 129, 55 \\
        & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.2e-6 & 134, 60 \\
        & g & Mende, East Africans & 60 & \textbf{2.13e-4} & 60 \\
        & h & East Africans, British & 50 & 4.16e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo & 12 &\textbf{3.37e-4} & 12 \\
        \textbf{Merger Model} & & & & & \\
        $LL= -101,600$ & a & Stem 1, Stem 2 & 1,692 & \textbf{1.26e-4} & 1,213 \\
        (Table~\ref{tab:supp-merger-with-stem-migration}) & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
        & b & Stem 2 to Nama & 119 & \textbf{0.70} & pulse \\
        & c & Stem 2 to Stem 1E & 98 & \textbf{0.52} & pulse \\
        & d & Stem 2 to Mende & 25 & \textbf{0.19} & pulse \\
        & e, f & Nama, Other Africans* & 119 & 4.5e-5, 9.8e-6 & 120, 60 \\
        & g & Mende, East Africans* & 60 & \textbf{1.97e-4} & 60 \\
        & h & East Africans, British & 50 & 3.82e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo &12 & \textbf{3.59e-4} & 12 \\
        \bottomrule
    \end{tabular}
    \caption{
        \textbf{Migration and divergence parameters from best fit models.}
        Labeled migration rates correspond to symmetric continuous migration
        bands shown in Figure~\ref{fig:best-fit-models}. Both the continuous migration and
        merger models inferred a relatively deep split of human stem branches,
        though they were connected by ongoing migration that maintained their
        genetic similarity. Bold text indicates migration rates above $10^{-4}$.
        In both models, the branch ancestral to the Nama shares a common ancestral population
        with the other African groups $\sim120-135$ka. Following this divergence,
        the population ancestral to other African groups branches into West and East African
        groups $60$ka.
        *Migration rates and durations are shown between branches ancestral to 
        1) Nama and East Africans and their ancestors, and
        2) Nama and Mende, respectively.
        ``Divergence times'' correspond to the most recent common ancestral population,
        and does not account for continuous migration or earlier reticulations.
    }
    \label{tab:migration-rates}
\end{table}

@@

To account for population structure prior to $135$ka, three of our four models
allowed for two or more ``stem'' populations which could diverge either before
or after the Neanderthal split. We considered models both with and without
migration between these stem populations, and in both cases we tested two
different types of gene exchange during the expansion phase, as illustrated in
Figure \ref{fig:mergers}: 1) One of the stem population expands (splits into
contemporary populations), and the other stem population(s) has continuous
symmetric migration with those populations; or 2) one or more of the stem
populations expands, with instantaneous pulse (or ``merger'') events from the
other stem population, so that recent populations are formed by mergers of
multiple ancestral populations. Depending on parameter values, this scenario
encompasses archaic hominin introgression and fragmentation-and-coalescence
models (such as Figure~\ref{fig:proposed-models}B and C).  For many parameters,
confidence intervals based on bootstrapping are relatively narrow
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. Model assumptions have,
however, a larger impact on parameter estimates (and thus, real uncertainty).
To convey model uncertainty, we highlight features of the two inferred models
with high likelihoods. These are referred to as the ``multiple-merger'' and the
``continuous-migration'' models. Both allow for migration between stem
branches, but differ primarily in the timing of the early divergence of stem
populations and their relative $N_e$ (Figure~\ref{fig:best-fit-models}). The
two models also differ in the mode of divergence, with the multiple-merger
featuring a population reticulation during the Middle Pleistocene.

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration between stems ($LL \approx
-101,600$ vs.\ $-107,700$ in the merger model, Tables
\ref{tab:supp-merger-without-stem-migration} and
\ref{tab:supp-merger-with-stem-migration}, and $LL \approx -115,300$ vs.\
$-126,500$ in the continuous migration model, Tables
\ref{tab:supp-continuous-migration-no-mig} and
\ref{tab:supp-continuous-migration-no-mig}).  With continuous migration between
stems, population structure extends back to up to 1.5Ma
(Table~\ref{tab:migration-rates}).  Migration between the stems in these models
is moderate, with a fraction of migrant lineages each generation estimated as
$m=6.3\times10^{-5}-1.3\times10^{-4}$. For comparison, this is similar to
inferred migration rates between connected contemporary populations over the
past $50$ka (Table~\ref{tab:migration-rates}). This ongoing (or at least,
periodic) gene flow qualitatively distinguishes these models from previously
proposed archaic hominin admixture models (Figure~\ref{fig:proposed-models}C)
as the early branches remain closely related, and each branch contributes large
amounts to all contemporary populations (Figure~\ref{fig:predictions}).
Because of this relatedness, only $1\%$ to $4\%$ of genetic differentiation
among contemporary populations can be traced back to this early population
structure (SI Section~\ref{sec:f4})

Under the continuous-migration model, one of the two stems diverges into
lineages leading to contemporary populations in western, southern and eastern
Africa, and the other (Stem 2) contributes variable ancestry to those
populations. This migration from Stem 2 is highest with the Mende
($m=1.6\times10^{-4}$) compared to the Nama and East African populations
($m=5.9\times10^{-5}$ and $3.1\times10^{-5}$, respectively), with migration
allowed to occur until $5$ka. A sampled lineage from the Nama, Mende, and Gumuz
have probabilities of being in Stem 2 at the time of Stem 1 expansion (135ka)
of approximately $0.145$, $0.2$, and $0.13$, respectively, though these
probabilities change over time, precluding the notion of a fixed admixture
proportion.

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups.  We observe a
sharp bottleneck in Stem 1 down to $N_e=100$ after the split of the Neanderthal
branch. This represents the lower bound allowed in our optimization (i.e. an
$N_e$ of 100), although the size of this bottleneck is poorly constrained (95\%
confidence interval 100--851). After a long period of exchange with Stem 2,
Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S'' 478ka. The timing of
this divergence was also poorly constrained ($95\%$ CI $276-478$ka).  These
populations evolve independently until $119$ka ($101-125$ka) when Stem 1S and Stem
2 combine to form the ancestors of the Nama, with proportions 30\%, 70\%
respectively. Similarly, Stem 1E and Stem 2 combine in equal proportions (50\%
each) to form the ancestors of the western Africans and eastern Africans (and
thus also all individuals who later disperse during the Out of Africa event).
Finally, the Mende receive a large additional pulse of gene flow from Stem 2,
replacing 19\% (18--21\%) of their population $25$ka ($22-26$ka). The later Stem 2
contribution to the western African Mende resulted in better model fits
($\Delta LL \approx 60,000$).  Speculatively, this may indicate that an
ancestral Stem 2 population occupied western or central Africa, broadly
speaking. The differing proportions in the Nama and eastern Africans may also
indicate geographic separation of Stem 1S in southern Africa and Stem 1E in
eastern Africa.

\begin{figure}[t!]
    \centering
    \includegraphics{figures/predictions.pdf}
    \caption{
        \textbf{Structure among stems is weak and present-day structure is mostly recent.}
        From the best fit models of our two parameterizations (A and B:
        continuous migration, C and D: merger with stem migration),
        we predicted pairwise differences $H_{i,j}$ between individuals
        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
        (B and D) To understand how drift between stems explains contemporary structure, 
        we also computed the proportion $\alpha^2$ of drift between pairs of
        sampled contemporary
        populations that aligns with drift between past populations
        (here Nama and Mende, see SI Section~S\ref{sec:f4}
        for details and additional comparisons in
        Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
        Both models infer deep population structure with modest contributions to
        contemporary genetic differentiation.
        Most present differentiation dates back to the last $100$ka.
    }
    \label{fig:predictions}
\end{figure}

@@

\begin{figure}[t!]
    \centering
    \includegraphics{figures/validation}
    \caption{
        \textbf{Model validation using independent statistics.} (A--C) Using
        our best fit models to linkage disequilibrium and pairwise diversity statistics,
        we simulated expected conditional site-frequency-spectra (cSFS) and
        compared to the observed cSFS from the data. Our inferred models
        provide a good fit to the data, even though these were not used in our
        inference. Across the three populations, ancestral state
        misidentification was consistently inferred to be $1.5-1.7\%$ for
        intergenic loci (SI \ref{sec:misspecification}). (D, E) We used Relate
        \citep{Speidel2019-nj} to reconstruct genome-wide gene genealogies,
        which we used to estimate coalescence rate trajectories
        and cross-coalescence rates between pairs of populations. While
        coalescence rate distributions are informative statistics about past
        evolutionary processes, interpretation can be hindered by migration and
        population structure, and translating relative cross-coalescence rate
        curves (RCCR) into population divergence times is especially prone
        to misinterpretation. For example, the Mende-Gumuz comparison shows
        a more recent increased RCCR than either population with the British,
        a pattern that is recapitulated under our best-fit model, even though
        the Mende-Gumuz split occurs prior to the Gumuz-British split.
    }
    \label{fig:validation}
\end{figure}

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time in real data and data simulated from our inferred
models. Many previous studies have found a reduction of coalescence rates
between $1000$ka and $100$ka in humans, and thus inferred an increase in $N_e$
during the same period \citep{Li2011-le}.  This increase in inferred $N_e$
could be due to either an increase in population size or to ancestral
population structure during the Middle Pleistocene \citep{Mazet2016-wn}.  All
models, including the single-origin model, recapitulate an inferred ancestral
increase in $N_e$ between $100$ka--$1$Ma extending deeper in the past than the
Relate estimate (Figure~\ref{fig:supp-iicr-sim} and SI
Section~\ref{inferred_relate}). Whereas the single-origin model achieves this by an
increase in $N_e$ during that period, the best-fit model recapitulate this
pattern without corresponding population size changes.

Relative cross-coalescence rates (rCCR) have recently been used to estimate
divergence between a pair of populations, as measured by the rate of
coalescence between two groups divided by the mean within population
coalescence. Simulations of rCCR accuracy, however, focus on a ‘clean split’
between populations whereby groups diverge without subsequent gene flow.
Published estimates of the earliest human divergences with rCCR, which range
from $150$ka--$100$ka \citep{Bergstrom2021-iw}, may be significantly biased when
compared to more complex models with gene flow as inferred here. We find that
midpoint estimates of rCCR are poor estimates for population divergence, often
underestimating divergence time by 50\% or greater (e.g., Mende vs.\ Gumuz
$\sim15$ka compared to a true divergence of $60$ka), and recent migration can
lead to the misordering of divergence events (Figure~\ref{fig:validation}E). We
suggest that rCCR analyses which do not fit multiple parameters including gene
flow should be interpreted with caution.
